---
title: "COVID-19 and R Part IV"
author: "Drew Tyre"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    code_folding: hide
draft: yes
slug: COVID-19_and_R_IV
tags_include:
- R
- OPD
- COVID19
categories: Research
---


# The bottom line

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library("tidyverse")
library("lubridate")

jhu_url <- paste("https://raw.githubusercontent.com/CSSEGISandData/", 
                 "COVID-19/master/csse_covid_19_data/", 
                 "csse_covid_19_time_series/", 
                 "time_series_19-covid-Confirmed.csv", sep = "")

#jhu_wide <- read_csv(jhu_url) # just grab it once from github
# archive it! 
#save(jhu_wide, file = "data/jhu_wide_2020-03-15.Rda")
load("data/jhu_wide_2020-03-15.Rda")
us_confirmed_total <- jhu_wide %>% 
  rename(province = "Province/State", 
         country_region = "Country/Region") %>% 
  pivot_longer(-c(province, country_region, Lat, Long), 
               names_to = "Date", values_to = "cumulative_cases") %>% 
  filter(country_region == "US") %>% 
  mutate(Date = mdy(Date)) %>% 
#  mutate(Date = lubridate::mdy(Date) - lubridate::days(1)) %>% 
  arrange(province, Date)  %>% 
  # filter out state rows prior to march 8, and county rows after that. 
  # this is dropping virgin islands ... 
  filter(str_detect(province, ", ") & Date <= "2020-03-9" |
           str_detect(province, ", ", negate = TRUE) & Date > "2020-03-9") %>% 
  group_by(Date) %>% # then group by Date and sum
  summarize(cumulative_cases = sum(cumulative_cases)) %>% 
  ungroup() %>% 
  mutate(incident_cases = c(0, diff(cumulative_cases)))

p1 <- ggplot(data = filter(us_confirmed_total, Date > "2020-02-28", Date <= "2020-03-11"),
       mapping = aes(x = Date)) + 
#  geom_line(mapping = aes(y = incident_cases)) + 
  geom_point(mapping = aes(y = incident_cases)) + 
  scale_y_log10() + 
  geom_smooth(mapping = aes(y = incident_cases), method = "lm")

us_exp_model <- us_confirmed_total %>% 
  mutate(log_incident_cases = log10(incident_cases),    # transform the data
         day = as.numeric(Date - ymd("2020-02-27"))) %>% # get day relative to Feb 27
  filter(Date > "2020-02-28", Date <= "2020-03-11") %>% 
  lm(data = .,
     formula = log_incident_cases ~ day) 

predicted <- tibble(day = 14:20)
predicted_list <- predict(us_exp_model, newdata = predicted, se.fit = TRUE)
predicted$fit <- predicted_list$fit # this is klutzy, but I want to see the answer! 
predicted$se.fit <- predicted_list$se.fit
predicted <- predicted %>% 
  mutate(Date = ymd("2020-2-27") + day,
         pred_var = se.fit^2 + predicted_list$residual.scale^2,
         lpl = 10^(fit - sqrt(pred_var)*qt(0.975, df = 11)),
         upl = 10^(fit + sqrt(pred_var)*qt(0.975, df = 11)),
         fit = 10^fit)

p1 + 
  geom_line(data = predicted,
            mapping = aes(y = fit),
            linetype = 2) +
  geom_ribbon(data = predicted,
            mapping = aes(ymin = lpl, ymax = upl),
            alpha = 0.2) +
  geom_point(data = filter(us_confirmed_total, Date > "2020-03-11"),
             mapping = aes(y = incident_cases)) +
  labs(y = "Incident cases", title = "Total new reported cases per day in the USA",
       subtitle = "Exponential model in blue; predicted values with 95% prediction intervals with dashed line.",
       caption = paste("Source data from https://github.com/CSSEGISandData/COVID-19 downloaded ",
                       format(file.mtime("data/jhu_wide_2020-03-15.Rda"), usetz = TRUE),
                       "\n Copyright Andrew Tyre 2020. Distributed with MIT License.")
       )
```


# Full data nerd below this point

[That article by Tomas Pueyo](https://medium.com/@tomaspueyo/coronavirus-act-today-or-people-will-die-f4d3d9cd99ca)
has some good points about the difference between reported cases (what I'm using) and true cases
(what I wish we had). The CDC does have a table of cases by date of infection, but it's a 
small fraction of the total cases reported. Nonetheless, I thought it might be 
instructive to run the exponential model on that data. Annnd, it's buried inside a JSON
file used by a javascript widget that makes the figure and displays the table. No HTML table to scrape. 

```{r, message=FALSE}
# if running this for the first time, uncomment (once!) all the lines from 
# <HERE> to 
# library("RJSONIO")
# library("RCurl")
# 
# # grab the data
# raw_data <- getURL("https://www.cdc.gov/coronavirus/2019-ncov/us-cases-epi-chart.json")
# # Then covert from JSON into a list in R
# data <- fromJSON(raw_data)$data$columns # the list item named "data" has what we want, in an item named columns
# # now put in a tibble, need to do some manual things to remove the name from the start of the data
# cdc_epi <- tibble(
#   Date = mdy(data[[1]][-1]), # grab first item in list, remove first element (the name)
#   incident_cases = as.numeric(data[[2]][-1]) # grab second item in list, remove first element
# )
# # archive it, so that I use the same data for these figures always!
# also means I don't hit the server with data requests every time I save the post.
# save(cdc_epi, file = "data/cdc_epi_2020-03-15.Rda")
# <HERE>
load(file = "data/cdc_epi_2020-03-15.Rda")

# plot it, just to see what it looks like
ggplot(data = cdc_epi,
       mapping = aes(x = Date, y = incident_cases)) + 
  geom_point() +
  geom_vline(xintercept = ymd("2020-03-05"), linetype = 2)
```

Cool! It worked ... OK, the vertical dashed line marks the point at which the CDC thinks
not all cases are yet reported, so we shouldn't model things after March 5th, yet.
I want to fit the same model as before, but the difficulty with that model is
that it doesn't handle zeros well -- $log_{10}(0) = -\infty$. In my 
current predictions I'm dodging that issue by only using data after
the number of incident cases is always positive. But I've been 
wondering how a glm with log link would do. Also need to choose an
error distribution; a Poisson distribution has only positive 
support and discrete values so I'll try that. 

```{r}
cdc_epi_trimmed <- filter(cdc_epi, Date < "2020-03-05") %>% 
  mutate(days = as.numeric(Date - ymd("2020-02-27"))) # use day relative to Feb 27, same as other model
cdc_epi_glm <- glm(incident_cases ~ days, data = cdc_epi_trimmed, family = poisson)
summary(cdc_epi_glm)
```
Wow. Daily growth rate is 0.14 day^-1^, almost exactly the same estimate as I get for the
model based on reported cases, because the base of the logarithms is different. This is 
easy to see by running the transformation based model on the cases by infection date data.
```{r}
cdc_epi_trimmed %>% 
  filter(Date > ymd("2020-02-17")) %>% 
  mutate(log_cases = log10(incident_cases)) %>% 
  lm(log_cases ~ days, data = .) %>% 
  summary()
```

So reported cases are growing faster than the number of cases with known infection dates. 
Trouble is, the number of cases with known infection dates is compromised by a lack of
reporting (or the numbers would match the reported cases). 

That model is mildly overdispersed
(ratio of residual deviance to degrees of freedom is 
`r format(cdc_epi_glm$deviance / cdc_epi_glm$df.residual, digits = 2)`), so
how does a Negative Binomial model compare?

```{r, message=FALSE}
library("mgcv") # switch to mgcv::gam to estimate overdispersion parameter
cdc_epi_glm2 <- gam(incident_cases ~ days, data = cdc_epi_trimmed, family = nb)
summary(cdc_epi_glm2)
```

The AIC for this model is `r format(AIC(cdc_epi_glm2),digits = 2)` so moderately better than
the Poisson model. 

## Fit the model for more countries

I'd like to get predictions for more countries so we can compare how the USA is doing with, say,
Italy or South Korea. I'm going back to reported case data for this, and I'm going to use the 
Negative Binomial model because then I don't have to worry about the odd zero close to the start of the time series. I'll process the other countries separately because of the funky county/state
business in the USA, then stick them together afterwards.

```{r}
countries <- c("Canada", "Australia", "Korea, South", "Italy", "Iran")
other_confirmed_total <- jhu_wide %>% 
  rename(province = "Province/State", 
         country_region = "Country/Region") %>% 
  pivot_longer(-c(province, country_region, Lat, Long), 
               names_to = "Date", values_to = "cumulative_cases") %>% 
  filter(country_region %in% countries,
         # have to trap the rows with missing province (most other countries)
         # otherwise str_detect(province ...) is missing and dropped by filter()
         is.na(province) | str_detect(province, "Princess", negate = TRUE)) %>% 
  mutate(Date = mdy(Date)) %>% 
  # filter out state rows prior to march 9, and county rows after that. 
  group_by(country_region, Date) %>% # then group by country and Date and sum
  summarize(cumulative_cases = sum(cumulative_cases)) %>% 
  group_by(country_region) %>% 
  mutate(incident_cases = c(0, diff(cumulative_cases))) %>% 
  ungroup()
us_confirmed_total <- mutate(us_confirmed_total, country_region = "USA")
all_confirmed_total <- bind_rows(other_confirmed_total, us_confirmed_total)

pall <- ggplot(data = all_confirmed_total,
       mapping = aes(x = Date)) + # don't add y = here, so we can change variables later for ribbons etc
  geom_point(mapping = aes(y = incident_cases)) + 
  facet_wrap(~country_region, scales = "free_y")
pall
```

Phew. That took far too long to sort out, because I tried to `filter()` all the countries including the
USA, and that was just not working. Treating the USA separately was easy because I already
had that dataframe built above. The y axis scales are wildly different on those, but just eyeballing 
the shape it's clear that South Korea is doing something radically different. There are also a couple
weird outliers in Australia and Italy where incident cases dropped to zero suddenly for one day. Canada
also has a strange 2 day oscillation going on. Now fit the models and see what that looks like.

```{r}
all_results <- all_confirmed_total %>%
  mutate(day = as.numeric(Date - ymd("2020-02-27"))) %>% # get day relative to Feb 27
  filter(Date > "2020-02-15", Date <= "2020-03-11") %>% 
  group_by(country_region) %>% 
  nest() %>% 
  mutate(models = map(data, ~gam(incident_cases ~ day, data = ., family = nb)))

```


[^1]: Code not shown in the post can be found [on Github](https://raw.githubusercontent.com/rbind/cheap_shots/master/content/post/covid-19_in_R_II.Rmd).