---
title: "How big is too big: variance inflation factors"
author: "Drew Tyre"
date: '2021-06-21'
output:
  html_document:
    code_folding: hide
  word_document: default
og_image: 
draft: no
slug: how-big-is-too-big-vif
tags_include:
- R
categories: Research
---

[In a comment Jacob E.](http://atyre2.github.io/2017/06/16/rebutting_cade.html) asked "Also, you mention using VIF for testing multi-collinearity. Is there a cutoff score you recommend for AIC?" I gave a snarky answer because I didn't have a good answer. What would a good answer look like?

One way to proceed is to simulate data from the situation where the assumption 
is not violated, and look at the distribution of VIF scores. 

```{r}
library("tidyverse")
n <- 100
N <- 1000
make_data <- function(beta = c(0, 0.1, 1, 10), sigma = 1, residual_error = 1, n = 1){
  if (is.null(dim(sigma))){
    tmp <- sigma
    sigma <- matrix(0, nrow = length(beta), ncol = length(beta))
    diag(sigma) <- tmp
  } 
  X <- MASS:::mvrnorm(n, mu = rep(0, length(beta)), Sigma = sigma)
  Y <- X %*% beta + rnorm(n, mean = 0, sd = residual_error)
  return(data.frame(Y, X))
}
#Test it out
test_df <- make_data(n = n)
ggplot(data = test_df,
       mapping = aes(x = X4, y = Y, color = X3)) + 
  geom_point()
```

OK, now generate a bunch of data sets, fit linear models to them, and get the
VIF scores.

```{r}
vif_null <- map(1:N, ~make_data(n = n)) %>% 
  map(~lm(Y~X1+X2+X3+X4, data = .)) %>% 
  map_dfr(car::vif) %>% 
  mutate(rep = 1:N) %>% 
  pivot_longer(cols = -rep, names_to = "variable",values_to = "vif")
ggplot(data = vif_null,
       mapping = aes(x = vif)) + 
  geom_histogram() + 
  facet_wrap(~variable)

```

OK, wow! Didn't expect that -- if there is NO correlation between the predictor 
variables then VIF rarely exceeds 1.2. Lets do an example where a couple of the
variables are highly correlated. 

```{r}
sigma <- matrix(c(1, 0, 0, 0,
                  0, 1, -0.9, 0,
                  0, -0.9, 1, 0,
                  0, 0, 0, 1), nrow = 4, ncol = 4)
vif_bad <- map(1:N, ~make_data(n = n, sigma = sigma)) %>% 
  map(~lm(Y~X1+X2+X3+X4, data = .)) %>% 
  map_dfr(car::vif) %>% 
  mutate(rep = 1:N) %>% 
  pivot_longer(cols = -rep, names_to = "variable",values_to = "vif")
ggplot(data = vif_bad,
       mapping = aes(x = vif)) + 
  geom_histogram() + 
  facet_wrap(~variable)
```

So a correlation of ~ -0.9 between two of the variables gives VIF scores bigger 
than 2, about 5 on average. So choosing a cutoff of 2 is going to exclude 
the worst offenders. 

