---
title: "What is the functional response of a recreational hunter?"
author: "Drew Tyre"
date: '2019-04-12'
output: 
  html_document:
    code_folding: hide
draft: yes
slug: hunter-functional-response
tags_include:
- ecology
- management
- modeling
categories: Research
---

A few years ago I built a white-tailed deer population model to evaluate different
harvest strategies. 2 sexes, density dependence, age structure, it was fun. The
main problem I came across was figuring out how changes in population
size and structure would interact with management to change harvest mortality. 
In consumer-resource modeling terminology, I didn't know what the functional 
response of hunters was. So here's my first shot at it.

To keep things simple, I will just think about 2 types of deer, does ($N_d$) 
and bucks ($N_b$). Furthermore, hunters always prefer bucks. In practice this
means if they encounter a buck they harvest it. The real trick is figuring out
what they do when they encounter a doe. Simplifying again, each permit only 
allows 1 deer, so if they take a doe then that precludes taking a buck. 
Each hunter has some limited time in which to harvest any animal (limited season). 
Simplifying, assume that's the same for all hunters. Finally, hunters encounter 
deer at rates proportional to their abundance $\lambda(N)$. This rate depends 
on deer behavior, hunter behavior, and the habitat involved.

If hunters have discrete intervals in which to encounter a deer, then this is kind
of like a negative binomial distribution -- how many non-bucks are encountered
before encountering a buck. The problem is we have two kinds of events, not one.
For a single interval

$$
p(buck | N) = 1-e^{\lambda N_b} \\
p(doe | N) = (1-e^{\lambda N_d})(1-p(buck)) \\
p(nothing) = (1 - p(buck))(1-p(doe))
$$

Here I assume that if a doe and a buck are encountered, the hunter always takes
the buck and ignores the doe. Also, if more than one buck or doe is encountered
that's the same as encountering one.

Maybe this is a dynamic programming problem -- there are T intervals, and on each
interval the decision is to take a doe or not, given you've encountered one. 

$$
V(t) = p(buck) + p(doe) max(v_d, V(t+1)) + p(nothing) V(t+1)
$$

where $0 \leq v_d \leq 1$ is the relative value of a doe compared to a buck.

```{r, message = FALSE, warnings = FALSE}
library("tidyverse")
lambda <-  0.01
N <-  c(doe = 100, buck = 20)
vd <-  0.5
p_enc <- function(N, lambda){
  c(pbuck = 1 - exp(-lambda*N[2]),
    pdoe = (1 - exp(-lambda*N[1]))*(exp(-lambda*N[2])),
    pnone = (exp(-lambda*N[1]))*(exp(-lambda*N[2])))
    }
V <- vector("numeric", length = 20)
V1 <- vector("numeric", length = 20)
V2 <- vector("numeric", length = 20)
D <- vector("character", length = 20)
p <- p_enc(N, lambda)
V[20] <- p[1] + vd*p[2]
V1[20] <- vd
V2[20] <- 0
D[20] <- "accept"

for(t in 19:1){
  V1[t] <- vd
  V2[t] <- V[t+1]
  if (V1[t] > V2[t]){
    D[t] <- "accept"
    V[t] <- V1[t]
  } else {
    D[t] <- "reject"
    V[t] <- V2[t]
  }
  V[t] <- p[1] + p[2] * V[t] + p[3] * V[t+1]
}
df <- bind_cols(t = 1:20, V = V, D = D, Accept = V1, Reject = V2) %>% 
  gather(key = Choice, value = Value, 4:5)
ggplot(data = df,
       mapping = aes(x = t)) +
  geom_line(mapping = aes(y = Value, linetype = Choice)) + 
  labs(x = "Time", y = "Value", 
       caption = "Expected value of rejecting or accepting a doe. N_doe = 100, N_buck = 20, lambda = 0.01, v_doe = 0.5")
```

 So with these parameters you should reject a doe until there are only a few
 opportunities left, and then you should start accepting them. That makes sense
 to me. Now I need to convert that into a model for the probability of harvesting
 a buck, doe, or nothing. 
 
 If I go back to my negative binomial thinking, if the hunter never accepts, the 
 future value of rejecting would be $1 - p(0 | T-t, p(buck))$. This is the 
 probability of getting at least one buck in $T-t$ opportunities. What does
 that look like?
 
```{r}
1-dnbinom(0,19:1, p[1])
```
 
That is much higher, clearly I haven't got this figured out yet. 

# As a function of $t_{switch}$

The DP tells me there will be a switch point. Before the switch, only accept bucks, 
after the switch, accept any deer. So I can write the value function like this

$$
V(t_{switch}) = (1 - (1-p(buck))^{t_{switch}}) + \\
(1-p(buck))^{t_{switch}}(1-p(none)^{T-t_{switch}})(p(doe)v_d + p(buck))
$$

The first term is the probability of getting a buck before the switch point. The 
second term is the probability of *not* getting a buck before the switch, times
the probability of getting *something*, times the expected value of something.

# Probability of harvesting a buck or doe

So this approach is getting to what I need, which is the probability a hunter
takes a buck or a doe, or is unsuccessful. Before the switch is easy; after the 
switch is harder.

$$
\begin{align}
P_{buck} = & (1 - (1-p(buck))^{t_{switch}}) + \\
  & (1-p(buck))^{t_{switch}}(p(buck) + p(none)p(buck) + ... + p(none)^{T-t_{switch}-1}p(buck)) \\
 = & (1 - (1-p(buck))^{t_{switch}}) + \\
  & (1-p(buck))^{t_{switch}}p(buck)(1-p(none)^{T-t_{switch}-1})
\end{align}
$$

and similarly for a doe, except we don't need the first term

$$
\begin{align}
P_{doe} = & (1-p(buck))^{t_{switch}}(p(doe) + p(none)p(doe) + ... + p(none)^{T-t_{switch}-1}p(doe)) \\
 = & (1-p(buck))^{t_{switch}}p(doe)(1-p(none)^{T-t_{switch}-1})
\end{align}
$$


