---
title: New blog title poll!
author: Drew Tyre
date: '2019-01-23'
slug: new-blog-title-poll
categories:
  - Personal
tags:
  - R
draft: no
image:
  caption: ''
  focal_point: ''
---

I'd like to change the title of this blog, so [I solicited suggestions on Twitter](https://twitter.com/atiretoo/status/1087407091252449283) and Facebook.
There were some great ones! Then I [held a poll](https://twitter.com/atiretoo/status/1087694051221024768).

```{r message=FALSE}
library("tidyverse")
poll <- tribble(
  ~Title, ~votes,
  "Irregularly Tyre-annical", 2,
  "Tyre Repair", 7,
  "Small things done great", 0,
  "Where the rubber meets the road", 6
)
ggplot(data = poll, aes(x=reorder(Title, votes), y = votes)) + 
  geom_col() +
  theme_minimal() + 
  coord_flip() + 
  labs(y = "Title", x = "Votes")
```

Boo. My personal contribution scored 0. Although the winner is "Tyre Repair"[^attributionTR], the
sample size is small, and I couldn't help wondering if there was a statistically
significant difference between that and "Where the rubber meets the road"[^attributionwtrmtr]. 

```{r}
library("EMT")
mt <- multinomial.test(poll$votes, prob = rep(1/4, 4))
mt$p.value
```

But that's not a terribly interesting hypothesis -- it says that my voters care
what title I use. What I really want to know 
is if there is any that beats all the others. I discovered [a post-hoc test for 
for the Exact GOF](http://www.biostathandbook.com/chigof.html), essentially
testing each category against the sum of the others, and using Bonferroni corrections.
So I'll have 4 comparisons, and thus my p value should be less than 0.05/4 = 0.0125. 

```{r}
pvals <- tribble(
  ~Comparison, ~p,
  "Irregularly Tyre-annical vs. others",  multinomial.test(c(2,13), prob = c(0.25, 0.75))$p.value,
  "Tyre Repair vs. others",  multinomial.test(c(7,8), prob = c(0.25, 0.75))$p.value,
  "Small things done great vs. others",  multinomial.test(c(0,15), prob = c(0.25, 0.75))$p.value,
  "Where the rubber meets the road vs. others",  multinomial.test(c(6,9), prob = c(0.25, 0.75))$p.value
)
```


[^attributionTR]: contributed by Poet extraordinaire Larkin Powell

[^attributionwtrmtr]: contributed by my stellar boss John Carroll 