---
title: New blog title poll!
author: Drew Tyre
date: '2019-01-23'
slug: new-blog-title-poll
categories:
  - Personal
tags:
  - R
draft: no
image:
  caption: ''
  focal_point: ''
---

I'd like to change the title of this blog, so [I solicited suggestions on Twitter](https://twitter.com/atiretoo/status/1087407091252449283) and Facebook.
There were some great ones! Then I [held a poll](https://twitter.com/atiretoo/status/1087694051221024768).

```{r message=FALSE}
library("tidyverse")
poll <- tribble(
  ~Title, ~votes,
  "Irregularly Tyre-annical", 2,
  "Tyre Repair", 7,
  "Small things done great", 0,
  "Where the rubber meets the road", 6
)
ggplot(data = poll, aes(x=reorder(Title,-votes), y = votes)) + 
  geom_col() +
  theme_minimal() + 
  coord_flip() + 
  labs(y = "Title", x = "Votes")
```

Boo. My personal contribution scored 0. Although the winner is "Tyre Repair"[^attributionTR], the
sample size is small, and I couldn't help wondering if there was a statistically
significant difference between that and "Where the rubber meets the road"[^attributionwtrmtr]. 

```{r}

```


[^attributionTR]: contributed by Poet extraordinaire Larkin Powell

[^attributionwtrmtr]: contributed by my stellar boss John Carroll 