---
title: "Rethinking the logistic model"
author: "Drew Tyre"
date: '2019-02-25'
output: html_document
draft: no
slug: rethink-logistic
tags_include:
- ecology
- carrying capacity
categories: Research
---

A few weeks ago I rambled on about [carrying capacity as a concept](https://drewtyre.rbind.io/posts/below-K/).
The day after I posted that, I stumbled across [a paper by James Mallet in 2012](https://mallet.oeb.harvard.edu/files/malletlab/files/mallet_the_struggle_2012.pdf)
on why the r-K parameterization of the logistic model is all wrong. Mind Blown. 

The essential point of the paper is that our typical form of logistic growth

$$
\frac{1}{N}\frac{dN}{dt} = r_{max}\left(1-\frac{N}{K}\right)
$$

is a historical accident. Verhulst's original form was

$$
\frac{1}{N}\frac{dN}{dt} = r_{max} - \alpha N
$$

with an equilibrium population size given by $N^* = r_{max} / \alpha$. This is brilliant, 
because now $r_{max}$ represents growth in the absence of intraspecific competition,
while $\alpha$ is how much growth is reduced by intraspecific competition. Most 
importantly, *these two parameters can vary independently*. In contrast, $r_{max}$ 
and $K$ are linked at the hip. 

To see this, consider the following diagram of per-capita birth and death rates changing with 
population size[^baseR].

```{r, echo=FALSE}
plot(0,0,type="n",xlab="Population Size",ylab="Per capita birth or death rates",axes=FALSE,xlim=c(0,1),ylim=c(0,1))
box(bty="l")
abline(a=0.9,b=-0.5,lwd=3)
abline(a=0.1,b=0.5,lwd=3,lty=3)
legend(0.8,1,bty="n",legend=c("Births","Deaths"),lty=c(1,3),lwd=3)
arrows(0,0.1,0,0.9,code=3)
text(0.01,0.5,"r",font=3)
arrows(0.8,0,0.8,0.5,code=2)
text(0.81,0,"K",font=3)
```

A change in per capita death rates, say by increasing trapping by humans that is 
density independent, shifts the death rate curve upwards, parallel to the current
line. This reduces $r_{max}$, because the y-intercepts get closer, *and reduces $K$*,
because now the two curves cross closer to the origin. However, $\alpha$, the sum
of the slopes of the lines *does not change*. 

This isn't a trivial side bar. Trevor Hefley's [work on detecting transcritical bifurcations
in real populations](https://rdcu.be/bnuBG) ran into issues because of this non-independence.
Trevor fit an r-K model to population data on Bobwhite. He allowed both $r_{max}$ and $K$ to 
change over time. If $r_{max}$ dips below 0 while $K > 0$, then the population 
goes through a "transcritical bifurcation" -- the equilibrium population size ceases
to exist, and the population decays exponentially towards 0. However, if $N>K$ when 
the bifurcation occurs, the r-K model predicts an increase towards $\infty$. 

```{r setup, message = FALSE, warning=FALSE}
library("tidyverse")
library("tidypop") # note: code below using the dev branch of this package use
# devtools::install_github("atyre2/tidypop@dev") 

rKmod <- function(N0, r, K){
  N1 <- N0 + N0*r*(1-N0/K)
  return(N1)
}
inputs <- tibble(t = 1:20,
                 r = c(rep(0.5,10), rep(-0.1,10)),
                 K = c(rep(100,10), rep(90,10)))
outputs <- iterate(inputs, N0 = 20, rKmod)
ggplot(data = outputs,
       mapping = aes(x = t, y = N)) + 
  geom_line() + 
  geom_step(data = inputs,
            mapping = aes(y = K), linetype = 2)
```

Needless to say that isn't the expected behavior for $r_{max} < 0$! If I switch
to the $r-\alpha$ form, I need to work out what $\alpha$ should be to have $N^* = 100$.
I think it is $r/N^* = 0.005$ in this case so 

```{r, warning=FALSE}
ramod <- function(N0, r, alpha){
  N1 <- N0 + N0*r - alpha*N0^2
  return(N1)
}
inputs <- tibble(t = 1:20,
                 r = c(rep(0.5,10), rep(-0.1,10)),
                 alpha = 0.005,
                 K = r / alpha)
outputs <- iterate(inputs, N0 = 20, rKmod)
ggplot(data = outputs,
       mapping = aes(x = t, y = N)) + 
  geom_line() + 
  geom_step(data = inputs,
            mapping = aes(y = K), linetype = 2) + 
  ylim(c(0,100))
```

That's more like it! The per capita effect of intra-specific competition didn't change -- 
the perturbation just affected the population in a density independent way. 

What is $\alpha$? I'll start with discrete time exponential growth, and substitute
straight line functions for the per capita birth and death rates.

$$
\begin{align}
N_{t+1} & =N_t\left(1 + b - d\right) \\
        & =N_t + N_tb - N_td \\
        & =N_t + N_t(b_0 + b_1N_t) - N_t(d_0 + d_1N_t) \\
        & =N_t + N_t(b_0 - d_0) + N_t^2(b_1 - d_1) \\
        & =N_t + r_{max}N_t -\alpha N_t^2
\end{align}
$$

So $b_0-d_0$ is usually rewritten as $r_{max}$. But also, $-\alpha = b_1 - d_1$,
which is ... just beautiful. To write that equation with $K$ requires so much more
algebra. 

It also expands easily to competition.
$$
\begin{align}
N_{1,t+1} & =N_{1,t} + r_{max}N_{1,t} -\alpha_{11} N_{1,t}^2 - \alpha_{21}N_{1,t}N_{2,t} \\
N_{2,t+1} & =N_{2,t} + r_{max}N_{2,t} -\alpha_{22} N_{2,t}^2 - \alpha_{12}N_{2,t}N_{1,t}
\end{align}
$$

and so on! Wow. 

Now I've even more reason to rewrite my lecture notes and redo all the videos. How do I check to see if
this change improves student learning? 

[^baseR]: Yes, that's using base R plotting. I'm being lazy and recycling a figure from my lecture notes, which are not yet tidied. 
